---
title: 'MEDI 504A: Working with Diabetes Data'
author: "Ehsan Karim & Liang Xu, ehsan.karim@ubc.ca"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document: default
  html_document: default
---

This document aims to present the general steps for analyzing binary data using machine learning methods. The data source is described in Strack B. et al. [1] [link](https://www.hindawi.com/journals/bmri/2014/781670/). The dataset can be downloaded from [here](https://archive.ics.uci.edu/ml/datasets/diabetes+130-us+hospitals+for+years+1999-2008). Below we present the codes for processing the analytic data following the guideline presented in the paper.

### Data Inspection
```{r, echo=FALSE, message=FALSE,warning=FALSE}
library(stringr)
library(ggplot2)
library(rpart) 
library(gbm)
require(pROC)
require(glmnet)
require(caret)
library(rms)
library(rpart)
library(data.table)
```


```{r, echo=FALSE, cache=TRUE, results='hide'}
require(data.table)
## read data
diabetic.data <- read.csv(file = "diabetic_data.csv", stringsAsFactors = TRUE)

## exclusions
# 1. only keep first obs from the same patient (with the same unique_id)
unique_id <- sapply(unique(diabetic.data$patient_nbr), FUN = function(x){
  which(diabetic.data$patient_nbr == x)[1]
})
diabetic.data = diabetic.data[unique_id, ]

# delete encounter_id now
diabetic.data$encounter_id <- NULL

# 2. remove patients that entered hospice or passed away after discharge
diabetic.data <- diabetic.data[ -which(diabetic.data$discharge_disposition_id %in% c(11, 13, 14, 19, 20, 21)), ]

# check missing data:
apply(diabetic.data, MARGIN = 2, FUN = function(x){length(which(x == "?"))})
# 'medical_specialty', 'weight' and 'payer_code' have lots of missing data

# remove the variables 'weight' and 'payer_code'
diabetic.data$weight <- NULL
diabetic.data$payer_code <- NULL

nrow(diabetic.data)
# 69973, which is 11 less than the value reported (69984)
```

### Coding of predictors

The predictors are coded following the steps outlined in the paper.
```{r, echo=FALSE, cache=TRUE, results='hide'}
# re-coding some of the variables:
diabetic.data$readmitted = as.factor(ifelse(diabetic.data$readmitted == "<30", "YES", "NO"))

diabetic.data$A1Cresult = as.factor(ifelse(diabetic.data$A1Cresult %in% c("Norm", ">7"), "Normal", ifelse(diabetic.data$A1Cresult == "None", "None", ifelse(diabetic.data$change == "Ch", "high_ch", "high_noch"))))

diabetic.data$discharge_disposition_id = as.factor(ifelse(diabetic.data$discharge_disposition_id == 1, "Home", "Other"))

diabetic.data$admission_source_id = as.factor(ifelse(diabetic.data$admission_source_id == 7, "Emergency", ifelse(diabetic.data$admission_source_id %in% c(1,2), "Referral", "Other")))

diabetic.data$medical_specialty <- replace(diabetic.data$medical_specialty, diabetic.data$medical_specialty == "Cardiology-Pediatric", "Cardiology")

diabetic.data$medical_specialty <- replace(diabetic.data$medical_specialty, diabetic.data$medical_specialty == "Surgeon", "SurgicalSpecialty")

diabetic.data$medical_specialty <- replace(diabetic.data$medical_specialty, diabetic.data$medical_specialty == "Surgery-Cardiovascular/Thoracic", "SurgicalSpecialty")
diabetic.data$medical_specialty <- replace(diabetic.data$medical_specialty, diabetic.data$medical_specialty == "Surgery-Colon&Rectal", "SurgicalSpecialty")
diabetic.data$medical_specialty <- replace(diabetic.data$medical_specialty, diabetic.data$medical_specialty == "Surgery-General", "SurgicalSpecialty")
diabetic.data$medical_specialty <- replace(diabetic.data$medical_specialty, diabetic.data$medical_specialty == "Surgery-Maxillofacial", "SurgicalSpecialty")
diabetic.data$medical_specialty <- replace(diabetic.data$medical_specialty, diabetic.data$medical_specialty == "Surgery-Neuro", "SurgicalSpecialty")
diabetic.data$medical_specialty <- replace(diabetic.data$medical_specialty, diabetic.data$medical_specialty == "Surgery-Pediatric", "SurgicalSpecialty")
diabetic.data$medical_specialty <- replace(diabetic.data$medical_specialty, diabetic.data$medical_specialty == "Surgery-Plastic", "SurgicalSpecialty")
diabetic.data$medical_specialty <- replace(diabetic.data$medical_specialty, diabetic.data$medical_specialty == "Surgery-PlasticwithinHeadandNeck", "SurgicalSpecialty")
diabetic.data$medical_specialty <- replace(diabetic.data$medical_specialty, diabetic.data$medical_specialty == "Surgery-Thoracic", "SurgicalSpecialty")
diabetic.data$medical_specialty <- replace(diabetic.data$medical_specialty, diabetic.data$medical_specialty == "Surgery-Vascular", "SurgicalSpecialty")
diabetic.data$medical_specialty <- str_replace(diabetic.data$medical_specialty, "SurgicalSpecialty", "Surgery")
diabetic.data$medical_specialty[which(diabetic.data$medical_specialty == "?")] = "Missing or Unknown" 
diabetic.data$medical_specialty[which(!(diabetic.data$medical_specialty %in% c("InternalMedicine", "Cardiology", "Surgery", "Family/GeneralPractice", "Missing or Unknown")))] = "Other" 
diag_diabetes = levels(diabetic.data$diag_1)[which(substr(levels(diabetic.data$diag_1), 1, 3)==250)]
diag_other = levels(diabetic.data$diag_1)[which(!(levels(diabetic.data$diag_1) %in% c(390:459, 785, diag_diabetes, 460:519, 786, 520:579, 787, 800:999, 710:739, 580:629, 788, 140:239)))]
levels(diabetic.data$diag_1) <- list(
  "Circulatory" = c(390:459, 785), 
  "Diabetes" = diag_diabetes, 
  "Respiratory" = c(460:519, 786), 
  "Digestive" = c(520:579, 787),
  "Injury and poisoning" = c(800:999),
  "Musculoskeletal" = c(710:739),
  "Genitourinary" = c(580:629, 788),
  "Neoplasms" = c(140:239),
  "Other" = diag_other
  )

levels(diabetic.data$race) = list("Missing" = "?", "Other" = c("Other", "Asian", "Hispanic"), "African American" = "AfricanAmerican", "Caucasian" = "Caucasian" )

diabetic.data$age_num = diabetic.data$age
levels(diabetic.data$age_num) = list("5" = "[0-10)", "15" = "[10-20)", "25" = "[20-30)", "35" = "[30-40)", "45" = "[40-50)", "55" = "[50-60)", "65" = "[60-70)", "75" = "[70-80)", "85" = "[80-90)", "95" = "[90-100)")

diabetic.data$age_num = as.numeric(as.character(diabetic.data$age_num))

levels(diabetic.data$age) = list("< 30" = c("[0-10)", "[10-20)", "[20-30)" ), "[30, 60)" = c("[30-40)", "[40-50)", "[50-60)"), "[60, 100) " = c("[60-70)", "[70-80)", "[80-90)", "[90-100)"))

diabetic.data$medical_specialty <- as.factor(diabetic.data$medical_specialty)
diabetic.data$medical_specialty <-droplevels(diabetic.data$medical_specialty)

# change the baseline level for some factors:
diabetic.data$race = relevel(diabetic.data$race, ref = "African American")

diabetic.data$age = relevel(diabetic.data$age, ref = "[30, 60)") 

diabetic.data$diag_1 = relevel(diabetic.data$diag_1, ref = "Diabetes") 

diabetic.data$A1Cresult =  relevel(diabetic.data$A1Cresult, ref = "None") 

# tables:
HbA1c = as.data.frame(cbind("HbA1c", levels(diabetic.data$A1Cresult), table(diabetic.data$A1Cresult),  (table(diabetic.data$A1Cresult))/nrow(diabetic.data), table(diabetic.data$A1Cresult, diabetic.data$readmitted)[, 2], table(diabetic.data$A1Cresult, diabetic.data$readmitted)[, 2]/table(diabetic.data$A1Cresult)))

gender = as.data.frame(cbind("gender", levels(diabetic.data$gender), table(diabetic.data$gender),  (table(diabetic.data$gender))/nrow(diabetic.data), table(diabetic.data$gender, diabetic.data$readmitted)[, 2], table(diabetic.data$gender, diabetic.data$readmitted)[, 2]/table(diabetic.data$gender)))

discharge = as.data.frame(cbind("discharge", levels(diabetic.data$discharge_disposition_id), table(diabetic.data$discharge_disposition_id),  (table(diabetic.data$discharge_disposition_id))/nrow(diabetic.data), table(diabetic.data$discharge_disposition_id, diabetic.data$readmitted)[, 2], table(diabetic.data$discharge_disposition_id, diabetic.data$readmitted)[, 2]/table(diabetic.data$discharge_disposition_id)))

admission = as.data.frame(cbind("admission", levels(diabetic.data$admission_source_id), table(diabetic.data$admission_source_id),  (table(diabetic.data$admission_source_id))/nrow(diabetic.data), table(diabetic.data$admission_source_id, diabetic.data$readmitted)[, 2], table(diabetic.data$admission_source_id, diabetic.data$readmitted)[, 2]/table(diabetic.data$admission_source_id)))

specialty = as.data.frame(cbind("specialty", levels(diabetic.data$medical_specialty), table(diabetic.data$medical_specialty),  (table(diabetic.data$medical_specialty))/nrow(diabetic.data), table(diabetic.data$medical_specialty, diabetic.data$readmitted)[, 2], table(diabetic.data$medical_specialty, diabetic.data$readmitted)[, 2]/table(diabetic.data$medical_specialty)))

diagnosis = as.data.frame(cbind("diagnosis", levels(diabetic.data$diag_1), table(diabetic.data$diag_1),  (table(diabetic.data$diag_1))/nrow(diabetic.data), table(diabetic.data$diag_1, diabetic.data$readmitted)[, 2], table(diabetic.data$diag_1, diabetic.data$readmitted)[, 2]/table(diabetic.data$diag_1)))

race = as.data.frame(cbind("race", levels(diabetic.data$race), table(diabetic.data$race),  (table(diabetic.data$race))/nrow(diabetic.data), table(diabetic.data$race, diabetic.data$readmitted)[, 2], table(diabetic.data$race, diabetic.data$readmitted)[, 2]/table(diabetic.data$race)))

age = as.data.frame(cbind("age", levels(diabetic.data$age), table(diabetic.data$age),  (table(diabetic.data$age))/nrow(diabetic.data), table(diabetic.data$age, diabetic.data$readmitted)[, 2], table(diabetic.data$age, diabetic.data$readmitted)[, 2]/table(diabetic.data$age)))

names(diabetic.data)[7] <- "source"
names(diabetic.data)[6] <- "discharge"
summarytable = rbind(HbA1c, gender, discharge, admission, specialty, diagnosis, race, age)
names(summarytable) = c("Variable", "Level", "# of encounters", "% of population", "# of encounters (readmitted)", "% in group (readmitted)")
row.names(summarytable) = NULL
summarytable$`% of population` = as.numeric(as.character(summarytable$`% of population`))
summarytable$`% in group (readmitted)` = as.numeric(as.character(summarytable$`% in group (readmitted)`))
summarytable$`% of population` = round((summarytable$`% of population`)* 100, digits = 2)
summarytable$`% in group (readmitted)` = round((summarytable$`% in group (readmitted)`)* 100, digits = 2)
summarytable

# mean and quantiles for numeric variables age and time in hospital:
mean(diabetic.data$age_num)
quantile(diabetic.data$age_num, probs = c(0.5, 0.25, 0.75))

mean(diabetic.data$time_in_hospital)
quantile(diabetic.data$time_in_hospital, probs = c(0.5, 0.25, 0.75))
```

### Model specifiction

For the purpose of comparison, we select the same set of predictors and interactions as in the paper. 

### Model Estimation 

Outcome and model specification:

```{r, cache=TRUE}
class(diabetic.data$readmitted)
levels(diabetic.data$readmitted)
model.formula <- as.formula("readmitted ~ discharge + race + source + medical_specialty + time_in_hospital + age + diag_1 + A1Cresult  +  diag_1 * discharge + race * discharge + medical_specialty*discharge + discharge*time_in_hospital + time_in_hospital*medical_specialty + age * medical_specialty + time_in_hospital*diag_1 + A1Cresult* diag_1")
```

1. Fit a logistic regression model using the above formula and the analytic data `diabetic.data`.

Hint: 
a) The results should be comparable to the values reported in Table 4 (but may not be exactly the same).
b) use `summary()` function to report the fit
c) No cross-validation required.

```{r}


lreg_fit <- glm(model.formula, data = diabetic.data, 
                family = binomial(link = "logit"))
summary(lreg_fit)

```

### Discrimination

To describe the discriminative ability of the model over different possible cutoffs, we can resort to the receiver operating characteristic (ROC) plot. The area under the ROC curve (AUC) is a popular indicator of how well the model performs with regards to discrimination. 

2. Report AUC from ROC.

```{r}
require(pROC)
dd.y2 <- diabetic.data$readmitted
prediction.y2 <- predict(lreg_fit, type = "response")
rocobject <- roc(dd.y2, prediction.y2)

rocobject 
plot(rocobject)
```

### Validation

Previously, we considered measures of performance using the whole dataset, and predictions of the same observations that were used to build the model. For a more realistic assessment of the model's performance, the model should be validated, and there are a couple of options: split-sample validation and K-fold cross-validation (CV). 

#### Cross-validation

3. Set up 10-fold cross-validation, fit logistic regression and obtain AUC from ROC from all the test datasets.

```{r, echo=FALSE}
set.seed(504)
require(caret)


ctrl<-trainControl(method = "cv", number = 10, 
                   classProbs = TRUE,
                   summaryFunction = twoClassSummary)


fit.cv.bin<-train(model.formula, trControl = ctrl,
               data = diabetic.data, method = "glm",
              family = binomial(),
              metric="ROC")
fit.cv.bin

summary.res <- fit.cv.bin$resample
summary.res
```

### Lasso, ridge or elastic net

4. Within 10-fold cross-validation, run the regularized regressions with the following parameter grids: `alpha = c(0,0.5,1), lambda = c(0.25, 0.75)`. Report the best alpha and lambda values the provides best AUC from ROC.  

```{r}
require(glmnet)



ctrl1<-trainControl(method = "cv", number = 10,
                   classProbs = TRUE,
                   summaryFunction = twoClassSummary)
fit.cv.bin1<-train(model.formula, trControl = ctrl1,
               data = diabetic.data, method = "glmnet",
               lambda= c(0.25, 0.75),
               tuneGrid = expand.grid(.alpha = c(0,0.5,1),  
                                      .lambda = c(0.25, 0.75)),
               verbose = FALSE,
               metric="ROC")
fit.cv.bin1
#best alpha value is 0, best lambda is 0.25, this is highest ROC which makes it a better model 
```

5. Plot the AUC from ROCs for all combinations of parameter grids used in the previous analysis.

```{r}
plot(fit.cv.bin1)
```


### Decision Trees

In addition to regression methods, the data can be explored with decision trees (specification of interaction not necessary). 

6. Within 10-fold cross-validation, run the regression trees. 

```{r}
require(caret)
set.seed(504)
model.formula0 <- as.formula("readmitted ~ discharge + race + source + medical_specialty + time_in_hospital + age + diag_1 + A1Cresult  +  time_in_hospital")
# your code here

ctrl<-trainControl(method = "cv", number = 10, 
                   classProbs = TRUE,
                   summaryFunction = twoClassSummary)

fit.cv.bin1<-train(model.formula0, trControl = ctrl,
               data = diabetic.data, method = "rpart",
              metric="ROC")
fit.cv.bin1

summary.res <- fit.cv.bin1$resample
summary.res
```


Variable importance: Report the 5 most important predictor categories. 
```{r}
# your code here
caret::varImp(fit.cv.bin1, scale = FALSE)
```

### Bagging

7. Within 10-fold cross-validation, run the bagging method. 

```{r}
set.seed(504)
require(caret)
ctrl<-trainControl(method = "cv", number = 10,
                   classProbs = TRUE,
                   summaryFunction = twoClassSummary)

fit.cv.bin2<-train(model.formula0, trControl = ctrl,
               data = diabetic.data, method = "bag",
               bagControl = bagControl(fit = ldaBag$fit, 
                                       predict = ldaBag$pred, 
                                       aggregate = ldaBag$aggregate),
               metric="ROC")

fit.cv.bin2

```


Variable importance: Report the 5 most important predictor categories.
```{r}
# your code here
caret::varImp(fit.cv.bin2, scale = FALSE)

#top 5 variables are discharge, time in hospital, age, medical specialty, source
```

### Boosting

8. Within 10-fold cross-validation, run the boosting method. 

```{r, results='hide',warning=FALSE,message=FALSE,error=FALSE}
# your code here
set.seed(504)
require(caret)
ctrl<-trainControl(method = "cv", number = 5,
                   classProbs = TRUE,
                   summaryFunction = twoClassSummary)

fit.cv.bin3<-train(model.formula0, trControl = ctrl,
               data = diabetic.data, method = "gbm",
               verbose = FALSE,
               metric="ROC")

plot(fit.cv.bin3)
fit.cv.bin3


```


I can't get fit.cv.bin3 to show up in the PDF:
Stochastic Gradient Boosting 

69973 samples
    8 predictor
    2 classes: 'NO', 'YES' 

No pre-processing
Resampling: Cross-Validated (5 fold) 
Summary of sample sizes: 55978, 55979, 55977, 55979, 55979 
Resampling results across tuning parameters:

  interaction.depth  n.trees  ROC        Sens  Spec
  1                   50      0.6033935  1     0   
  1                  100      0.6047011  1     0   
  1                  150      0.6051064  1     0   
  2                   50      0.6047578  1     0   
  2                  100      0.6062765  1     0   
  2                  150      0.6068501  1     0   
  3                   50      0.6045371  1     0   
  3                  100      0.6061718  1     0   
  3                  150      0.6063597  1     0   

Tuning parameter 'shrinkage' was held constant at a value of 0.1
Tuning parameter 'n.minobsinnode'
 was held constant at a value of 10
ROC was used to select the optimal model using the largest value.
The final values used for the model were n.trees = 150, interaction.depth = 2, shrinkage = 0.1
 and n.minobsinnode = 10.

Variable importance: Report the 5 most important predictor categories.
```{r}
caret::varImp(fit.cv.bin3, scale = FALSE)


```


[1] Beata Strack, Jonathan P. DeShazo, Chris Gennings, Juan L. Olmo, Sebastian Ventura, Krzysztof J. Cios, John N. Clore, "Impact of HbA1c Measurement on Hospital Readmission Rates: Analysis of 70,000 Clinical Database Patient Records", BioMed Research International, vol. 2014, Article ID 781670, 11 pages, 2014. https://doi.org/10.1155/2014/781670